### 2017-06-01 Manuscript/Response_to_reviewers.Rmd

\bigskip

**L 165: “time point t + 1” Is the time span between repeated samples always the same? If not, please add a sentence to make it clear. Are samples taken in the same month every year?**

RESPONSE: Samples are taken as and when we can catch the birds, which is not always in the same month. We now clarify this on lines XX:

\textit{"Note that the follow-up time between samples varied among individuals, and that individuals are not always caught at exactly the same time of year."}

@@

---

\bigskip

**Some more information on the sampled individuals your study is based on would be helpful and could be added to the supplement. E.g. cohort sizes.**

RESPONSE: We have now added a table of cohort sizes to the supplement (Table S1), and refer to this in the main text on Line XX.

@@

---

\bigskip

**L 168: How do you explain a non-normal distribution of telomere lengths? Data selection or methodological effect? How large is the variation of samples among cohorts?**

RESPONSE: Positive skew in telomere length distribution is a widespread and uncontroversial phenomenon, and the causes are likely biological (see for example Golubev et al. 2003 J. Theor. Biol.; Hirt et al. 2014; J. Math. Biol.; ). Investigating the causes of this are outside the scope of our study, and we feel that a discussion of this would be an unnecessary distraction.

### 2017-06-01 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

For a large subset of birds we had longitudinal data, with two or more samples taken at different ages (n = `r length(unique(c(dd3$BloodID,dd3$BloodID1)))` measurements from `r length(unique(dd3$BirdID))` birds). For these individuals we calculated the within-individual change in RTL by subtracting RTL at time point *t* from RTL at time point *t* + 1 (hereafter $\Delta$RTL, n = `r nrow(dd3)` measurements). Negative values of RTL reflect decreases in telomere length with age, while positive values reflect increases.

@@

##Results
We measured telomere lengths using a total of `r length(unique(dd$BloodID))` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`. Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were `r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))` and `r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))` respectively. Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice at different time points, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`). Using adult samples greater than one year old, we checked whether RTL was related to sample storage time, and found no evidence of such a relationship (estimate = `r round3(fixef(RTL_ST)['StorageTime'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_ST)['StorageTime',],lessthan = F),collapse = ', ')`).

@@

In addition to age, RTL was associated with tarsus length, sex and insect abundance (Fig. 3A). RTL was negatively related to tarsus length and males had longer telomeres than females (Fig. 3B), while insect abundance was positively related to RTL (Fig. 3C). The full model was weak in terms of explanatory power of fixed effects (marginal R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2m'])`), although including the random effect terms increased this substantially (conditional R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2c'])`). The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig. S2). One interesting finding from the model selection was that sex only appeared in top models where tarsus length was also present (Table S1). In accordance with this, when tarsus length was removed from the full model sex was no longer significant (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`), and a sex x tarsus interaction was significant when included (estimate = `r round3(fixef(RTL_Full_SexTarsus)['SexMales:Tarsus'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_SexTarsus)['SexMales:Tarsus',],lessthan = F),collapse = ', ')`); RTL decreased with tarsus length in both sexes, but this decrease was stronger in females (Fig. 3B). No social or ecological environmental variables were significant predictors of $\Delta$RTL using the full model approach (Table S2). Using model selection, we found that the top model explaining $\Delta$RTL contained age and population density (Table S3). $\Delta$RTL was positively related to age, consistent with telomere shortening being highest in early life, and negatively related to population density; however, in both instances model averaged confidence intervals overlapped zero (Fig. S3).

>>>>

For a large subset of birds we had longitudinal data, with two or more samples taken at different ages (n = `r length(unique(c(dd3$BloodID,dd3$BloodID1)))` measurements from `r length(unique(dd3$BirdID))` birds). For these individuals we calculated the within-individual change in RTL by subtracting RTL at time point *t* from RTL at time point *t* + 1 (hereafter $\Delta$RTL, n = `r nrow(dd3)` measurements). Negative values of RTL reflect decreases in telomere length with age, while positive values reflect increases. Note that the follow-up time between samples varied among individuals, and that individuals are not always caught at exactly the same time of year.

@@

##Results
We measured telomere lengths using a total of `r length(unique(dd$BloodID))` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)` (Table S1). Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were `r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))` and `r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))` respectively. Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice at different time points, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`). Using adult samples greater than one year old, we checked whether RTL was related to sample storage time, and found no evidence of such a relationship (estimate = `r round3(fixef(RTL_ST)['StorageTime'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_ST)['StorageTime',],lessthan = F),collapse = ', ')`).

@@

In addition to age, RTL was associated with tarsus length, sex and insect abundance (Fig. 3A). RTL was negatively related to tarsus length and males had longer telomeres than females (Fig. 3B), while insect abundance was positively related to RTL (Fig. 3C). The full model was weak in terms of explanatory power of fixed effects (marginal R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2m'])`), although including the random effect terms increased this substantially (conditional R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2c'])`). The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S2; Fig. S2). One interesting finding from the model selection was that sex only appeared in top models where tarsus length was also present (Table S2). In accordance with this, when tarsus length was removed from the full model sex was no longer significant (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`), and a sex x tarsus interaction was significant when included (estimate = `r round3(fixef(RTL_Full_SexTarsus)['SexMales:Tarsus'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_SexTarsus)['SexMales:Tarsus',],lessthan = F),collapse = ', ')`); RTL decreased with tarsus length in both sexes, but this decrease was stronger in females (Fig. 3B). No social or ecological environmental variables were significant predictors of $\Delta$RTL using the full model approach (Table S3). Using model selection, we found that the top model explaining $\Delta$RTL contained age and population density (Table S4). $\Delta$RTL was positively related to age, consistent with telomere shortening being highest in early life, and negatively related to population density; however, in both instances model averaged confidence intervals overlapped zero (Fig. S3).

### 2017-06-01 Manuscript/SW_early_life_telomeres_SOM.Rmd
Edits

**Table S1** Model selection table for analysis of factors affecting early life telomere length in juvenile Seychelles warblers. Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.

```{r Table S1}

@@

tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
rownames(tableS1) <- NULL
tableS1$Model[tableS1$Model == ''] <- 'Null model'
tableS1[,3:5] <- lapply(tableS1[,3:5],as.numeric)
kable(tableS1,digits=2)

@@

**Table S2** Full model of factors affecting $\Delta$RTL in Seychelles warblers, using longitudinal data.

@@

**Table S3** Model selection table for analysis of factors affecting $\Delta$RTL in Seychelles warblers, using longitudinal data. Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here. TQ = territory quality.

@@

tableS3 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
colnames(tableS3) <- c('Model','df','AICc', 'Delta AICc','Weight')
rownames(tableS3) <- NULL
tableS3$Model[tableS3$Model == ''] <- 'Null model'
tableS3[,3:5] <- lapply(tableS3[,3:5],as.numeric)
kable(tableS3,digits=2)

>>>>

**Table S1** Cohort sizes for each birth year of the Seychelles warbler dataset used in this study.
```{r Table S1}
x <- subset(dd,!duplicated(BirdID)) %>%
  ddply(.(LayYear),
        summarise,
        N = length(RTL)) %>%
  `colnames<-`(c("Cohort", "N"))
kable(x)
```

**Table S2** Model selection table for analysis of factors affecting early life telomere length in juvenile Seychelles warblers. Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.

@@

tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
colnames(tableS2) <- c('Model','df','AICc', 'Delta AICc','Weight')
rownames(tableS2) <- NULL
tableS2$Model[tableS2$Model == ''] <- 'Null model'
tableS2[,3:5] <- lapply(tableS2[,3:5],as.numeric)
kable(tableS2,digits=2)

@@

**Table S3** Full model of factors affecting $\Delta$RTL in Seychelles warblers, using longitudinal data.

@@

**Table S4** Model selection table for analysis of factors affecting $\Delta$RTL in Seychelles warblers, using longitudinal data. Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here. TQ = territory quality.

```{r Table S4}

@@

tableS4 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
colnames(tableS4) <- c('Model','df','AICc', 'Delta AICc','Weight')
rownames(tableS4) <- NULL
tableS4$Model[tableS4$Model == ''] <- 'Null model'
tableS4[,3:5] <- lapply(tableS4[,3:5],as.numeric)
kable(tableS4,digits=2)
